\section{Conclusion}
\label{sec:conclusion}

We have presented MorphoCLIP, a contrastive learning framework that maps Cell Painting images and natural-language perturbation descriptions into a shared 512-dimensional space. The design brings together three capabilities that prior work has only paired in twos -- text supervision, optional cross-well batch alignment, and joint training over both small molecules and genetic perturbations -- and trains on a single 16\,GB GPU thanks to a pre-extraction pipeline that caches the frozen-backbone outputs. About 14M parameters are trained, against 1.48B total in the closest text-based baseline.

The empirical results are preliminary and mixed. On the CPJUMP1 validation split, text-to-image retrieval reaches Recall@10 of 24.3\% with a median rank of 28, roughly $2.3\times$ a like-for-like random baseline; image-to-text retrieval, by contrast, is close to chance in aggregate. These numbers come from a single seed, a text-only and batch-correction-off configuration, and a well-level evaluation protocol, so they bound what the current setup shows rather than establish the method.

If a perturbation-level evaluation and the planned ablations confirm that the signal holds up, the use cases this design points toward include searching for existing drugs whose cellular signature matches a target phenotype, suggesting a likely mechanism of action from an unknown compound's neighbors in embedding space, and querying image databases in plain English without engineering bespoke features.

\subsection{Limitations and Future Work}

The most immediate limitation is the evaluation protocol. We rank individual wells, where each perturbation carries about 24 replicate positives; the standard CPJUMP1 protocol instead aggregates replicates into one profile per perturbation and ranks over the perturbation set. Re-running the evaluation at the perturbation level is the first priority, both to compare cleanly against prior work and because aggregation should reduce the image-side noise that leaves image-to-text retrieval near chance today.

The second limitation is that the design's central claim is asserted rather than measured. Cross-well alignment is disabled in the reported runs, and the factorial ablation in Table~\ref{tab:ablation_design} that would attribute performance to text supervision, batch correction, and gene-inclusive training has not yet been run. Until it lands -- across the three seeds and the held-out test split -- the contribution of each component is a hypothesis. The prompt-richness ablation will separately test whether SMILES strings and gene-function text add signal through BioClinical ModernBERT or are inert.

Beyond closing those gaps, three directions are worth pursuing. First, scale: CPJUMP1 is the pilot, and the full JUMP corpus contains $\sim$140{,}000 perturbations across many labs, microscopes, and cell lines, which is the right setting to test generalization. Second, richer prompts: pulling pathway and mechanism annotations from UniProt, Reactome, and DrugBank should strengthen the text encoder's biological grounding. Third, zero-shot transfer: running the trained encoders on Cell Painting datasets from labs and cell lines not seen during training would show whether the learned representations are general or specific to CPJUMP1.

We plan to release model weights, the evaluation suite, and an interactive natural-language retrieval demo on publication.
