\section{Introduction}

Phenotypic drug discovery aims to identify how a chemical or genetic perturbation changes the state of a cell, and to do so at a scale that target-specific assays cannot reach. This matters because the path from candidate to approved drug costs on the order of \$2\,B and takes about a decade~\cite{wouters2020rdcost}, and roughly nine of ten compounds that enter clinical trials fail~\cite{sun2022clinicaltrials}.

High-content microscopy offers a scalable alternative. The Cell Painting assay~\cite{bray2016cellpainting, seal2024cellpainting} stains live cells with five fluorescent dyes covering DNA, endoplasmic reticulum, RNA/nucleoli, actin/Golgi, and mitochondria, producing a multi-channel image of cellular state. Morphological changes after perturbation encode the perturbation's mechanism. The CPJUMP1 benchmark~\cite{chandrasekaran2024jump} casts this as a retrieval problem: given $\sim$3 million field-of-view images covering 303 compounds, 160 CRISPR knockouts, and 176 ORF overexpressions, can a model match perturbations that act on the same biological target?

Existing methods make uneven progress. The classical CellProfiler pipeline~\cite{carpenter2006cellprofiler} extracts handcrafted shape, texture, and intensity features and recovers only 5--25\% of expected compound--gene matches on CPJUMP1~\cite{chandrasekaran2024jump}. Recent contrastive methods do better, but each covers a different part of the problem. CLOOME~\cite{sanchez2023cloome} pairs cell images with Morgan fingerprints and reaches $70\times$ random retrieval, but only for compounds. MolPhenix~\cite{fradkin2024molphenix} improves on this with a pre-trained Phenom-1 encoder, again restricted to compounds. CellCLIP~\cite{lu2025cellclip} replaces fingerprints with text prompts and a CrossChannelFormer over DINOv2-g (1.48B parameters), enabling all three perturbation types through language, but its cross-class mAP still trails CellProfiler. CWA-MSN~\cite{huang2025cwamsn} instead targets batch effects directly, using cross-well alignment for correction; with only 22M parameters it improves on CellCLIP by 9\% on gene--gene retrieval, but it has no text branch and so cannot be queried in natural language.

We propose \textbf{MorphoCLIP}, a contrastive learning framework that brings text supervision, optional cross-well alignment for batch correction, and joint training over both small molecules and genetic perturbations into a single model. Following CLIP~\cite{radford2021clip}, MorphoCLIP aligns Cell Painting images and natural-language perturbation descriptions in a shared 512-dimensional space. Our working hypothesis is that text supervision and batch correction are complementary: language adds semantic structure that purely visual methods cannot recover, while cross-well alignment removes technical variation that otherwise swamps fine-grained biological signal. Whether the two combine as hoped is a question we frame here and leave to the planned ablations, since the results in this paper are from a text-only, batch-correction-off configuration.

The contributions of this paper are:
\begin{enumerate}
    \item A parameter-efficient contrastive framework built on a frozen DINOv3 ViT-L/16 backbone~\cite{simeoni2025dinov3} ($\sim$300M frozen) and a CrossChannelFormer aggregator, with about 14M trainable parameters against CellCLIP's 1.48B total.
    \item A single model trained jointly on compounds, CRISPR knockouts, and ORF overexpressions through a shared text encoder, where prior contrastive methods are limited to compounds.
    \item A design that supports optional cross-well alignment and a factorial ablation to attribute performance to text supervision, batch correction, and gene-inclusive training.
    \item A preliminary CPJUMP1 evaluation reported against a like-for-like random baseline, in which text-to-image retrieval reaches Recall@10 of 24.3\% ($\sim$2.3$\times$ random) while image-to-text retrieval stays close to chance.
\end{enumerate}
