\section{Related Work}

\subsection{Morphological Profiling}

Cell Painting~\cite{bray2016cellpainting} introduced the now-standard five-stain protocol for high-content morphological profiling and has become the dominant assay for high-throughput phenotypic screening~\cite{seal2024cellpainting}. The classical analysis tool, CellProfiler~\cite{carpenter2006cellprofiler}, extracts more than 1,000 handcrafted shape, texture, and intensity features per cell. On CPJUMP1~\cite{chandrasekaran2024jump}, this pipeline retrieves 5--25\% of expected sister perturbations within a modality and 7--17\% of CRISPR guide matches; cross-modal compound-to-gene matching is harder still.

Handcrafted features are bounded by the measurements their authors chose to compute and miss higher-order phenotypic patterns. Self-supervised vision transformers trained with DINO surpass CellProfiler on drug-target and gene-family classification with zero-shot transfer~\cite{moshkov2025ssl}, which motivates the use of frozen foundation-model features for Cell Painting.

\subsection{Contrastive Learning for Cell Painting}

CLOOME~\cite{sanchez2023cloome} was the first CLIP-style framework for Cell Painting. It jointly trains a ResNet image encoder and a Morgan-fingerprint MLP and reaches top-1 retrieval $>70\times$ random over $\sim$2,000 candidates. CLOOME handles only chemical perturbations and treats the five channels as RGB, discarding channel-specific signal.

MolPhenix~\cite{fradkin2024molphenix} replaces the trained image branch with a frozen Phenom-1 encoder, averages replicate embeddings, and introduces a soft-to-label (S2L) loss, gaining $8.78\times$ over CLOOME. It also covers compounds only and depends on a proprietary backbone.

CellCLIP~\cite{lu2025cellclip} is the closest prior work. It contributes (i)~templated text prompts (``A Cell Painting image of [cell\_type] cells treated with [drug\_name], SMILES: [SMILES\_string]''), (ii)~a CrossChannelFormer that processes each channel separately before cross-attention, (iii)~attention-based MIL pooling across cells, and (iv)~the Continuously Weighted Contrastive Loss (CWCL) for soft positives. CellCLIP runs on a DINOv2-g backbone (1.1B parameters, 1.48B total) and is evaluated on the Bray dataset~\cite{ljosa2012annotated} and on RxRx3-core gene--gene benchmarks (CORUM recall 0.354). Although CellCLIP supports all perturbation types through text, its cross-class mAP falls below the CellProfiler baseline, suggesting that text supervision on its own does not solve the batch-effect problem.

\subsection{Batch-Effect Correction}

CWA-MSN~\cite{huang2025cwamsn} attacks batch effects directly with cross-well alignment inside a masked siamese network. By aligning embeddings of cells subjected to the same perturbation across different wells and plates, it absorbs technical variance without an explicit batch label. With 22M parameters and 0.2M training images, CWA-MSN improves on CellCLIP by 9\% and on OpenPhenom by 29\% on gene--gene retrieval (CORUM recall 0.386). It has no text branch.

\subsection{Foundation Models for Microscopy}

DINOv2~\cite{oquab2024dinov2} and DINOv3~\cite{simeoni2025dinov3} provide self-supervised visual representations that transfer well to biological imaging despite being trained on natural images~\cite{moshkov2025ssl}. DINOv3 adds gram anchoring and an improved training recipe over DINOv2. Channel Vision Transformers~\cite{bao2024channelvit} show that per-channel processing followed by cross-channel attention is effective for multi-channel microscopy, which motivates our CrossChannelFormer aggregator.

On the text side, BioClinical ModernBERT~\cite{sounack2025bioclinical} (150M parameters) is pre-trained on 53.5B tokens of PubMed, PMC, and clinical text, roughly $17\times$ the biomedical corpus used by PubMedBERT (110M, 3.1B tokens) in CellCLIP. Built on ModernBERT with RoPE, Flash Attention, and GeGLU activations, it can encode both small-molecule and gene descriptions in a single model, unlike SMILES-specialized encoders such as ChemBERTa.

\subsection{Positioning of MorphoCLIP}

Table~\ref{tab:landscape} places MorphoCLIP relative to these methods. The gap we target is the absence of any prior method that combines text supervision with both gene-inclusive training and batch-effect correction. CellCLIP has text but no gene training and no batch correction; CWA-MSN has genes and batch correction but no text. MorphoCLIP fills that intersection with a parameter-efficient architecture ($\sim$8M trainable parameters versus CellCLIP's 1.48B total).

\begin{table}[t]
\centering
\caption{Landscape of Cell Painting representation learning methods. MorphoCLIP uniquely combines text supervision, gene-inclusive training, and batch correction.}
\label{tab:landscape}
\resizebox{\columnwidth}{!}{%
\begin{tabular}{@{}lccccc@{}}
\toprule
Method & Image Encoder & Pert.\ Types & Batch Corr.\ & Params \\
\midrule
CellProfiler~\cite{carpenter2006cellprofiler} & Handcrafted & All & Sphering & N/A \\
CLOOME~\cite{sanchez2023cloome} & ResNet & Compounds & No & ${\sim}$25M \\
MolPhenix~\cite{fradkin2024molphenix} & Phenom-1 & Compounds & No & ${\sim}$36M \\
CellCLIP~\cite{lu2025cellclip} & DINOv2-g & All (text) & No & 1,477M \\
CWA-MSN~\cite{huang2025cwamsn} & ViT-S/16 & Unimodal & CWA & 22M \\
\textbf{MorphoCLIP} & DINOv3-L & All (text) & CWA & ${\sim}$8M \\
\bottomrule
\end{tabular}%
}
\end{table}
