\section{Results and Discussion}

\subsection{Cross-Modal Retrieval Performance}

Table~\ref{tab:retrieval} reports retrieval performance of MorphoCLIP on the CPJUMP1 validation split alongside the published CellCLIP and CellProfiler numbers. The baselines were measured on different datasets and task formulations, so the cross-method comparisons are indicative; we discuss the caveats below.

\begin{table}[t]
\centering
\caption{Cross-modal retrieval. MorphoCLIP is evaluated on CPJUMP1 (2{,}220 wells, 817 candidate perturbations). CellCLIP numbers are taken from the original paper, evaluated on the Bray dataset. CellProfiler uses profile-based sister-perturbation matching on CPJUMP1.}
\label{tab:retrieval}
\begin{tabular}{@{}llrrr@{}}
\toprule
Model & Direction & R@1 & R@5 & R@10 \\
\midrule
CellProfiler & sister matching & \multicolumn{3}{c}{FR: 4.3--25.1\%} \\
\midrule
\multirow{2}{*}{CellCLIP} & perturb $\to$ profile & 1.18 & 4.49 & 7.37 \\
 & profile $\to$ perturb & 1.25 & 4.82 & 7.39 \\
\midrule
\multirow{2}{*}{\textbf{MorphoCLIP}} & image $\to$ text & 5.72 & 6.71 & 7.07 \\
 & text $\to$ image & 3.51 & \textbf{14.86} & \textbf{24.32} \\
\bottomrule
\end{tabular}
\end{table}

The headline number is text-to-image Recall@10 of 24.3\%, two orders of magnitude above the random baseline of 0.12\% on the same 817-candidate pool, with a median rank of 28 (top 3.4\%). Image-to-text retrieval is substantially harder: Recall@10 is 7.07\% with a median rank of 289 (still below the random baseline of 408, but only by a factor of about $1.4\times$).

\subsection{Retrieval Asymmetry}

The two directions are not symmetric: text-to-image R@10 is 24.3\% versus image-to-text R@10 of 7.1\%, with median ranks of 28 and 289 respectively. Most of this gap is structural rather than algorithmic. The validation split has many wells per perturbation, so a text query has multiple valid image positives to retrieve, while an image query must identify a single correct label out of 817. CellCLIP avoids this asymmetry because it operates on aggregated profiles (one vector per perturbation) rather than individual well images, making image-to-text and text-to-image largely equivalent on its setup.

\subsection{Comparison with Baselines}

\paragraph{Versus CellCLIP.} On text-to-image retrieval, MorphoCLIP reaches R@10 = 24.3\% on a candidate pool of 817 perturbations, against CellCLIP's best reported R@10 of 7.4\% on its evaluation set. The two evaluations are not strictly comparable: the candidate pools differ in size and composition, and CellCLIP retrieves over aggregated profiles while MorphoCLIP retrieves over individual well images, which favors the text-to-image direction (see above). On the more directly comparable image-to-text direction, MorphoCLIP R@10 of 7.07\% is essentially tied with CellCLIP's profile-to-perturb R@10 of 7.39\%. The contribution we want to highlight is therefore not raw R@10 but parameter efficiency: $\sim$8M trainable parameters versus CellCLIP's 1.48B total, with comparable image-to-text performance and joint training over all three perturbation types instead of compounds only.

\paragraph{Versus CellProfiler.} On CPJUMP1, CellProfiler's fraction retrieved ranges from 4.3\% to 25.1\% (mean $\sim$13\%) depending on the modality and matching task~\cite{chandrasekaran2024jump}. MorphoCLIP's text-to-image R@10 sits above the upper end of that range, on the easier text-query direction. The qualitative difference, independent of any number, is that MorphoCLIP supports natural-language querying of an image database, which CellProfiler does not.

\paragraph{Versus CWA-MSN.} A direct head-to-head comparison with CWA-MSN requires running both methods on the same gene--gene retrieval benchmarks (CORUM, HuMAP, Reactome). CWA-MSN's CORUM recall of 0.386 with 22M parameters is the relevant target. The factorial ablation in Table~\ref{tab:ablation_design} is designed to answer the more specific question of whether text supervision adds anything over CWA alone, which is the place we expect MorphoCLIP to provide a contribution.

\subsection{Feature Analysis}

We inspect the pre-extracted DINOv3 features to check that channel structure survives the frozen backbone. On plate BR00116991 (500 sites), a variance decomposition attributes 31.1\% of feature variance to differences between channels, 37.1\% to differences between sites, and 31.8\% to residual variation. The between-channel component is large enough to suggest that the per-channel encoding does carry channel-specific signal, even though DINOv3 was trained on natural images. A PCA over the $5 \times 1024$-dimensional feature space reaches 90\% of variance with 23 principal components, indicating moderate intrinsic dimensionality.

Cross-site same-channel similarity (mean 0.84) is higher than cross-channel similarity at the same site (mean 0.75), a gap of 0.09 that holds for every channel pair. The DNA/nucleus channel is the most distinct, with a mean similarity of 0.73 to the other four channels, consistent with its very different biological content.

\subsection{Computational Efficiency}

Pre-extraction is what makes MorphoCLIP cheap to iterate on (Table~\ref{tab:efficiency}). The DINOv3 forward pass is the dominant cost; once features are cached, the contrastive training loop is small enough to fit comfortably on a single 16\,GB GPU.

\begin{table}[t]
\centering
\caption{Wall-clock profile of the training loop on a single RTX 5080 (16\,GB VRAM). The LoRA row is an indicative estimate from a partial run, included to motivate the pre-extraction design rather than as a measured baseline.}
\label{tab:efficiency}
\begin{tabular}{@{}lrrr@{}}
\toprule
Configuration & Train.\ Params & VRAM & Time/epoch \\
\midrule
Pre-extracted + CCF (ours) & $\sim$8M & $\sim$2\,GB & 3--5 min \\
Frozen DINOv3 + LoRA ($r{=}8$, est.) & $\sim$20M & $\sim$14\,GB & 45--90 min \\
\bottomrule
\end{tabular}
\end{table}

Feature extraction is a one-time cost of about 7 minutes per plate ($\sim$6 hours for all 51). After that, an epoch over the cached features takes 3--5 minutes wall-clock, which is what enables the ablation study to run dozens of configurations on a single GPU in a day. Both DINOv3 (300M parameters) and BioClinical ModernBERT (150M parameters) stay frozen throughout; only the CrossChannelFormer and the two projection heads are updated.
